### 2018-03-07 chapter2/ms.tex
Chapter 2 cross-referencing first pass

To estimate the expected number of events observed during a GW optical follow-up search we compute the number of sources per square degree per year. This is done by determining the maximum redshift to which the object can be observed to a limiting magnitude of $z \approx$ 23 mag (see Sections~\ref{sec:ch2_kilo} and~\ref{sec:ch2_MCsims_det}). This rate is then multiplied by the search area $(A \sim 100$ deg$^2$) and duration to get the number of expected sources for our observations. We compute the duration by assuming that a kilonova will be detected within approximately one day of the GW trigger. Therefore, we do not consider events that begin more than +2 d into the search. However, we do consider sources that begin up to twice their characteristic timescale ($\sim5$ d for rapid transients and $\sim 20$ d for Type Ia SNe) prior to the GW trigger, as they may remain visible when the search begins. The total duration is then 7 d for rapid transients and 22 d for Type Ia SNe. The number of expected sources is given in Tables~\ref{tab:ch2_rates} and~\ref{tab:ch2_rates_Iae}. Type Ia SNe dominate the number of expected events with $N \sim 568$ when summing across all redshift bins. Of the population of fast transients, only type .Ia and the Pan-STARRS fast transients are expected to have $N \gtrsim 1$. Despite the low expected occurrence rate for other fast transients, we still consider these sources in this analysis due to the large uncertainty in their rates.

@@

Using these results we determine the fraction of sources that do not exhibit a sufficient change in brightness to be identified as a transient after the second or third observation. We begin by considering the first and second columns of Figures~\ref{fig:ch2_delm2} and~\ref{fig:ch2_delm3} (Strategies 1 and 2). Recall that for these strategies the time between the first and second epoch is $\sim$1 day. In both of these cases, the rapid evolution of the {\em r}-band light curve means that there is always a detectable change in brightness, but the source will always be seen in decline. Furthermore, it is still the case that given the overall faintness of kilonovae in {\em r}-band, only $\sim$13\% of sources will be detected at $r\approx24$ mag.

@@

The rapid decline of $\sim0.6-1.3$ mag in {\em r}-band that is evident in Figures~\ref{fig:ch2_delm2} and~\ref{fig:ch2_delm3} hints at a possible detection strategy. Namely, the GW error region can be observed with a cadence of $\sim$1 d, with rapidly declining sources flagged. However, a depth of $\approx$26 mag is required for a high detection fraction even if we require only 2 detections (to separate kilonovae from stellar flares), so covering the large GW error region can only be achieved with LSST.

@@

\begin{figure}[t!]
\centering
\includegraphics[width=0.9\textwidth]{./figs/chapter2/ch2_f10.pdf}
\caption{The decline time-color slice of the decline time-color-magnitude phase-space. The plot is constructed using the same methodology as in Figures~\ref{fig:ch2_mcphasecol} and~\ref{fig:ch2_mcphasedm}. We note that, as before, kilonovae show redder $i-z$ colors than the contaminant population. However, there is no strong separation in decline time.}
\label{fig:ch2_deccol}
\end{figure}

@@

Turning to the contaminant population, we find that several of the source classes have poorly constrained colors with some events exhibiting $i-z > 0$ mag. We can assess the impact of these contaminants by utilizing the rate data in Tables~\ref{tab:ch2_rates} and~\ref{tab:ch2_rates_Iae}. This is achieved by multiplying the expected fraction of sources in a region of the phase-space with the number of sources expected to occur during the search. This gives the actual number of sources detected. As expected, Type Ia SNe are the most numerous contaminant with $\sim3-12$ sources expected to exhibit $i-z\gtrsim0$ mag. Strategies 3 and 4 produce the fewest number of Type Ia SNe detections ($\sim 3-5$), while Strategy 2 produces the most ($\sim 12$). For the Pan-STARRS fast transients, we find that the number of sources exhibiting $i-z\gtrsim0$ mag is $\sim3$ across all four strategies. Lastly, we expect $\ll1$ type .Ia and miscellaneous fast transients to exhibit $i-z\gtrsim0$ mag. If we additionally require that the rise time is $\lesssim 4$ days the number of detected Type Ia SNe is reduced to $\sim1-7$. This does not affect the number of detected Pan-STARRS fast transients. Ultimately, we do not expect the lack of a pre-existing template to significantly hinder our ability to separate kilonovae from contaminants on the basis of their $i-z$ color and timescale, if the source is observed during the rise to peak.

@@

\begin{figure}[h!]
\centering
\includegraphics[width=0.9\textwidth]{./figs/chapter2/ch2_f18.pdf}
\caption{{\em r}-band light curves from the alternative kilonova models discussed in Sections~\ref{sec:ch2_HMNS} and ~\ref{sec:ch2_neutronpre}. The {\em r}-band light curve for the neutron-rich dynamical ejecta with $M_{{\rm ej}} = 10^{-2}\;M_{\odot}$ and $\beta_{{\rm ej}} = 0.2$ is plotted for comparison. The alternative models produce light curves which can be significantly brighter and bluer than the dynamical ejecta but peak at earlier times.}
\label{fig:ch2_altLC}
\end{figure}

\subsection{Implications for {\em r}-band detectability}
\label{sec:ch2_altkilo_det}
We now investigate the effect of these speculative models on the early-time detectability of kilonovae using {\em r}-band observations. We construct the light curves for the long-lived HMNS from the time-resolved spectra modeled by \citet{Kasen+15} using the procedure outlined in \cref{sec:ch2_kilo}. These spectra are for a HMNS lifetime of 100 ms with a dynamical torus with $10^{-2}\;M_{\odot}$ of neutron-rich material observed at a viewing angle of $\cos{\theta} = 0.95\; (\theta \approx 18$ deg). For the neutron precursor, we use the {\em r}-band light curve provided by \citet{Metzger+15}, for $10^{-4}\;M_{\odot}$ of free neutrons with an electron fraction of $Y_e \approx 0.05$ and opacity of $\kappa \approx 30$ cm$^2$ g$^{-1}$. These light curves are plotted in \cref{fig:ch2_altLC} along with those from the dynamical ejecta-only model for $M_{{\rm ej}} = 10^{-2}\;M_{\odot}$ and $\beta_{{\rm ej}} = 0.2$ used in Sections~\ref{sec:ch2_MCsims} and ~\ref{sec:ch2_diff}.

@@

The first and second points are the most important considerations as the sizable localization regions from the early science-runs of Advanced LIGO present the largest observational challenge and have the most pronounced impact on the quality of data that can be obtained. The key issue here is whether or not there will be sufficient time to observe the entire GW localization region on a nightly basis. A successful GW follow-up campaign must be able to observe a significant fraction of the GW localization region in a single night to obtain the optimal cadence. The third question impacts the likelihood that we will be able to observe a kilonovae before peak. This is important as we have shown in Sections~\ref{sec:ch2_MCsims_cont} and~\ref{sec:ch2_diff} that it is easier to isolate a kilonova from the contaminant population if it is observed during the rise. Additionally, observing a kilonova at early times is important both for constraining the ejecta parameters and for testing the alternative kilonovae models outlined in \cref{sec:ch2_altkilo}. Ultimately, our ability to address the three questions above depends heavily on the facilities in use.

@@

\begin{deluxetable}{cccccc}
\tabletypesize{\footnotesize}
\tablecolumns{6}
\tabcolsep0.05in\footnotesize
\tablewidth{0pt}
\tablecaption{Expected Rates For Fast Transients}
\tablehead{
\colhead{Object}&  \colhead{$\mathcal{R}_{\rm vol}$}& \colhead{$z_{\rm max}$}  &  \colhead{$\mathcal{R}_{\rm area}$} &  \colhead{$N$}     &     \colhead{Reference}      \\
\colhead{} & \colhead{(Mpc$^{-3}$ yr$^{-1}$)} & \colhead{} & \colhead{(deg$^{-2}$ yr$^{-1}$)} &  \colhead{} & \colhead{} }
\startdata
 Type .Ia    &                  10$^{-6}$                   &  0.5        &  $0.7$            &         1.4  & \citet{Bildsten+07} \\
 WD-NS mergers  &                  10$^{-5}$                   &   0.06        & $2\times10^{-2}$              &         $3\times10^{-2}$  &     \citet{Thompson2009}     \\
 WD-BH mergers  &      10$^{-5}$                   &    0.06        &   $2\times10^{-2}$              &      $3\times10^{-2}$  &    \citet{Fryer+99}          \\
AIC       &                  10$^{-6}$                   & 0.06        & $2\times10^{-3}$            &        $3\times10^{-3}$ &  \citet{Darbha+10}   \\
ELDD      &                  3$\times 10^{-7}$                   & 0.14        &    $6\times10^{-3}$           &      $1\times10^{-2}$ &  $\cdots$  \\
Pan-STARRS fast &                  5$\times 10^{-6}$                   &     0.5        &   $3.5$            &     7 &   \citet{Drout+14}
\enddata
\tablecomments{Expected rates for the various contaminants considered in \cref{sec:ch2_contaminants}. $\mathcal{R}_{\rm area}$ is computed assuming an isotropic distribution of sources in a volume defined by the comoving volume at $z_{\rm max}$.  The column $N$ refers to the number of events expected during a search covering 100 deg$^2$ for 7 days. See \S~\ref{sec:ch2_rates} or details.}
\label{tab:ch2_rates}
\end{deluxetable}

\begin{deluxetable}{cccc}
\tabletypesize{\footnotesize}
\tablecolumns{4}
\tabcolsep0.05in\footnotesize
\tablewidth{0pt}
\tablecaption{Expected Rates For Type Ia SNe}
\tablehead{
\colhead{Redshift}     &  \colhead{$\mathcal{R}_{\rm vol}$}&  \colhead{$\mathcal{R}_{\rm area}$}  &     \colhead{$N$} \\
\colhead{} & \colhead{(Mpc$^{-3}$ yr$^{-1}$)} &  \colhead{(deg$^{-2}$ yr$^{-1}$)} &  \colhead{} }
\startdata
0 -- 0.25 & $3\times10^{-5}$ & 3.5 & 21  \\
0.25 -- 0.50 & $4\times10^{-5}$ & 25 & 153  \\
0.5 -- 0.75 & $5\times10^{-5}$ & 65 & 394
\enddata
\tablecomments{Expected rates for Type Ia SNe in redshift bins of $\Delta z = 0.25$ out to $z_{\rm max} = 0.75$. See \S~\ref{sec:ch2_rates}, \citet{Graur+14}, and references therein.}
\label{tab:ch2_rates_Iae}
\end{deluxetable}

>>>>

To estimate the expected number of events observed during a GW optical follow-up search we compute the number of sources per square degree per year. This is done by determining the maximum redshift to which the object can be observed to a limiting magnitude of $z \approx$ 23 mag (see \cref{sec:ch2_kilo,sec:ch2_MCsims_det}). This rate is then multiplied by the search area $(A \sim 100$ deg$^2$) and duration to get the number of expected sources for our observations. We compute the duration by assuming that a kilonova will be detected within approximately one day of the GW trigger. Therefore, we do not consider events that begin more than +2 d into the search. However, we do consider sources that begin up to twice their characteristic timescale ($\sim5$ d for rapid transients and $\sim 20$ d for Type Ia SNe) prior to the GW trigger, as they may remain visible when the search begins. The total duration is then 7 d for rapid transients and 22 d for Type Ia SNe. The number of expected sources is given in \cref{tab:ch2_rates,tab:ch2_rates_Iae}. Type Ia SNe dominate the number of expected events with $N \sim 568$ when summing across all redshift bins. Of the population of fast transients, only type .Ia and the Pan-STARRS fast transients are expected to have $N \gtrsim 1$. Despite the low expected occurrence rate for other fast transients, we still consider these sources in this analysis due to the large uncertainty in their rates.

@@

Using these results we determine the fraction of sources that do not exhibit a sufficient change in brightness to be identified as a transient after the second or third observation. We begin by considering the first and second columns of \cref{fig:ch2_delm2,fig:ch2_delm3} (Strategies 1 and 2). Recall that for these strategies the time between the first and second epoch is $\sim$1 day. In both of these cases, the rapid evolution of the {\em r}-band light curve means that there is always a detectable change in brightness, but the source will always be seen in decline. Furthermore, it is still the case that given the overall faintness of kilonovae in {\em r}-band, only $\sim$13\% of sources will be detected at $r\approx24$ mag.

@@

The rapid decline of $\sim0.6-1.3$ mag in {\em r}-band that is evident in \cref{fig:ch2_delm2,fig:ch2_delm3} hints at a possible detection strategy. Namely, the GW error region can be observed with a cadence of $\sim$1 d, with rapidly declining sources flagged. However, a depth of $\approx$26 mag is required for a high detection fraction even if we require only 2 detections (to separate kilonovae from stellar flares), so covering the large GW error region can only be achieved with LSST.

@@

\begin{figure}[t!]
\centering
\includegraphics[width=0.9\textwidth]{./figs/chapter2/ch2_f10.pdf}
\caption{The decline time-color slice of the decline time-color-magnitude phase-space. The plot is constructed using the same methodology as in \cref{fig:ch2_mcphasecol,fig:ch2_mcphasedm}. We note that, as before, kilonovae show redder $i-z$ colors than the contaminant population. However, there is no strong separation in decline time.}
\label{fig:ch2_deccol}
\end{figure}

@@

Turning to the contaminant population, we find that several of the source classes have poorly constrained colors with some events exhibiting $i-z > 0$ mag. We can assess the impact of these contaminants by utilizing the rate data in \cref{tab:ch2_rates,tab:ch2_rates_Iae}. This is achieved by multiplying the expected fraction of sources in a region of the phase-space with the number of sources expected to occur during the search. This gives the actual number of sources detected. As expected, Type Ia SNe are the most numerous contaminant with $\sim3-12$ sources expected to exhibit $i-z\gtrsim0$ mag. Strategies 3 and 4 produce the fewest number of Type Ia SNe detections ($\sim 3-5$), while Strategy 2 produces the most ($\sim 12$). For the Pan-STARRS fast transients, we find that the number of sources exhibiting $i-z\gtrsim0$ mag is $\sim3$ across all four strategies. Lastly, we expect $\ll1$ type .Ia and miscellaneous fast transients to exhibit $i-z\gtrsim0$ mag. If we additionally require that the rise time is $\lesssim 4$ days the number of detected Type Ia SNe is reduced to $\sim1-7$. This does not affect the number of detected Pan-STARRS fast transients. Ultimately, we do not expect the lack of a pre-existing template to significantly hinder our ability to separate kilonovae from contaminants on the basis of their $i-z$ color and timescale, if the source is observed during the rise to peak.

@@

\begin{figure}[h!]
\centering
\includegraphics[width=0.9\textwidth]{./figs/chapter2/ch2_f18.pdf}
\caption{{\em r}-band light curves from the alternative kilonova models discussed in \cref{sec:ch2_HMNS,sec:ch2_neutronpre}. The {\em r}-band light curve for the neutron-rich dynamical ejecta with $M_{{\rm ej}} = 10^{-2}\;M_{\odot}$ and $\beta_{{\rm ej}} = 0.2$ is plotted for comparison. The alternative models produce light curves which can be significantly brighter and bluer than the dynamical ejecta but peak at earlier times.}
\label{fig:ch2_altLC}
\end{figure}

\subsection{Implications for {\em r}-band detectability}
\label{sec:ch2_altkilo_det}
We now investigate the effect of these speculative models on the early-time detectability of kilonovae using {\em r}-band observations. We construct the light curves for the long-lived HMNS from the time-resolved spectra modeled by \citet{Kasen+15} using the procedure outlined in \cref{sec:ch2_kilo}. These spectra are for a HMNS lifetime of 100 ms with a dynamical torus with $10^{-2}\;M_{\odot}$ of neutron-rich material observed at a viewing angle of $\cos{\theta} = 0.95\; (\theta \approx 18$ deg). For the neutron precursor, we use the {\em r}-band light curve provided by \citet{Metzger+15}, for $10^{-4}\;M_{\odot}$ of free neutrons with an electron fraction of $Y_e \approx 0.05$ and opacity of $\kappa \approx 30$ cm$^2$ g$^{-1}$. These light curves are plotted in \cref{fig:ch2_altLC} along with those from the dynamical ejecta-only model for $M_{{\rm ej}} = 10^{-2}\;M_{\odot}$ and $\beta_{{\rm ej}} = 0.2$ used in \cref{sec:ch2_MCsims,sec:ch2_diff}.

@@

The first and second points are the most important considerations as the sizable localization regions from the early science-runs of Advanced LIGO present the largest observational challenge and have the most pronounced impact on the quality of data that can be obtained. The key issue here is whether or not there will be sufficient time to observe the entire GW localization region on a nightly basis. A successful GW follow-up campaign must be able to observe a significant fraction of the GW localization region in a single night to obtain the optimal cadence. The third question impacts the likelihood that we will be able to observe a kilonovae before peak. This is important as we have shown in \cref{sec:ch2_MCsims_cont,sec:ch2_diff} that it is easier to isolate a kilonova from the contaminant population if it is observed during the rise. Additionally, observing a kilonova at early times is important both for constraining the ejecta parameters and for testing the alternative kilonovae models outlined in \cref{sec:ch2_altkilo}. Ultimately, our ability to address the three questions above depends heavily on the facilities in use.

@@

\begin{deluxetable}{cccccc}
\tabletypesize{\footnotesize}
\tablecolumns{6}
\tabcolsep0.05in\footnotesize
\tablewidth{0pt}
\tablecaption{Expected Rates For Fast Transients}
\tablehead{
\colhead{Object}&  \colhead{$\mathcal{R}_{\rm vol}$}& \colhead{$z_{\rm max}$}  &  \colhead{$\mathcal{R}_{\rm area}$} &  \colhead{$N$}     &     \colhead{Reference}      \\
\colhead{} & \colhead{(Mpc$^{-3}$ yr$^{-1}$)} & \colhead{} & \colhead{(deg$^{-2}$ yr$^{-1}$)} &  \colhead{} & \colhead{} }
\startdata
 Type .Ia    &                  10$^{-6}$                   &  0.5        &  $0.7$            &         1.4  & \citet{Bildsten+07} \\
 WD-NS mergers  &                  10$^{-5}$                   &   0.06        & $2\times10^{-2}$              &         $3\times10^{-2}$  &     \citet{Thompson2009}     \\
 WD-BH mergers  &      10$^{-5}$                   &    0.06        &   $2\times10^{-2}$              &      $3\times10^{-2}$  &    \citet{Fryer+99}          \\
AIC       &                  10$^{-6}$                   & 0.06        & $2\times10^{-3}$            &        $3\times10^{-3}$ &  \citet{Darbha+10}   \\
ELDD      &                  3$\times 10^{-7}$                   & 0.14        &    $6\times10^{-3}$           &      $1\times10^{-2}$ &  $\cdots$  \\
Pan-STARRS fast &                  5$\times 10^{-6}$                   &     0.5        &   $3.5$            &     7 &   \citet{Drout+14}
\enddata
\tablecomments{Expected rates for the various contaminants considered in \cref{sec:ch2_contaminants}. $\mathcal{R}_{\rm area}$ is computed assuming an isotropic distribution of sources in a volume defined by the comoving volume at $z_{\rm max}$.  The column $N$ refers to the number of events expected during a search covering 100 deg$^2$ for 7 days. See \cref{sec:ch2_rates} or details.}
\label{tab:ch2_rates}
\end{deluxetable}

\begin{deluxetable}{cccc}
\tabletypesize{\footnotesize}
\tablecolumns{4}
\tabcolsep0.05in\footnotesize
\tablewidth{0pt}
\tablecaption{Expected Rates For Type Ia SNe}
\tablehead{
\colhead{Redshift}     &  \colhead{$\mathcal{R}_{\rm vol}$}&  \colhead{$\mathcal{R}_{\rm area}$}  &     \colhead{$N$} \\
\colhead{} & \colhead{(Mpc$^{-3}$ yr$^{-1}$)} &  \colhead{(deg$^{-2}$ yr$^{-1}$)} &  \colhead{} }
\startdata
0 -- 0.25 & $3\times10^{-5}$ & 3.5 & 21  \\
0.25 -- 0.50 & $4\times10^{-5}$ & 25 & 153  \\
0.5 -- 0.75 & $5\times10^{-5}$ & 65 & 394
\enddata
\tablecomments{Expected rates for Type Ia SNe in redshift bins of $\Delta z = 0.25$ out to $z_{\rm max} = 0.75$. See \cref{sec:ch2_rates}, \citet{Graur+14}, and references therein.}
\label{tab:ch2_rates_Iae}
\end{deluxetable}
